---
title: "Answering the Question: In what ways do members and casual riders use Divvy bikes differently?"
author: "Chi Kit Yeung"
# output: html_notebook
output: html_document
---
<!-- ## Guiding questions -->

<!-- - Where is your data located? -->
<!-- - How is the data organized? -->
<!-- - Are there issues with bias or credibility in this data? Does your data ROCCC? -->
<!-- - How are you addressing licensing, privacy, security, and accessibility? -->
<!-- - How did you verify the data’s integrity? -->
<!-- - How does it help you answer your question? -->
<!-- - Are there any problems with the data? -->

---

## Setting up the environment

Loading all the necessary packages

```{r setup}
library(tidyverse)
library(janitor)
library(lubridate)
library(hms)
library(knitr)
```

## Loading the Data

The data is stored in csv files that are organized monthly. 

```{r}
# Loading the data
path <- "./divvy-tripdata/"
filenames <- list.files(path = path)
```

### Into a list

Since our data is being store in multiple csv files, I'll be using a loop to store each data frame into a list.  Doing so avoids cluttering up local environment.

```{r}
all_df <- lapply(filenames, function(i) {
  i <- paste(path,i,sep="")
  read.csv(i, header=FALSE)
  })
filenames <- gsub("-","_",filenames)
names(all_df) <- gsub(".csv","",filenames)
```

### Wrangling the Data


```{r}
joined_df <- all_df %>% 
  reduce(full_join)

# Renaming the data frame
## reduce() function renamed the columns with some place holder values and placed the actual names into the first row of the data frame.
names(joined_df) <- joined_df[1,]
df <- joined_df %>% 
  slice(-1)
```

#### Let's see what it's like

```{r}
str(df)
```

### Cleaning the Data

Looks like all the data is imported as 'character' data type.  Converting them to the correct types.

```{r}
## Converting to correct data types, removing entries with NA
df_clean <- df %>% 
  mutate(started_at = ymd_hms(started_at)) %>% 
  mutate(ended_at = ymd_hms(ended_at)) %>% 
  mutate(start_lat = as.double(start_lat)) %>% 
  mutate(start_lng = as.double(start_lng)) %>%
  mutate(end_lat = as.double(end_lat)) %>%
  mutate(end_lng = as.double(end_lng)) %>%
  mutate(ride_length = as_hms(difftime(ended_at, started_at)), .after = ended_at) %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  na.omit() %>% 
  filter(started_at<ended_at) %>% # remove cases where end time is somehow before or equals to the start time
  as_tibble()
```


#### Calculating some basic stats

```{r}
## Getting average amount of time different rider types ride
df_stats <- df_clean %>% 
  group_by(member_casual) %>% 
  summarise(mean = as.duration(round(mean(ride_length))), 
            max = as.duration(round(max(ride_length))), 
            min = as.duration(round(min(ride_length))),
            sd = as.duration(round(sd(ride_length))),
            median = as.duration(round(median(ride_length)))
            )

  kable(df_stats)
```

Wow.. the longest ride is _5 weeks_?! There's definitely some outliers here.  Probably someone who took out a bike and forgot to return it (o_o")

---

#### Applying some statistics to help filter out the outliers

```{r}
df_quantile <- df_clean %>% 
  group_by(member_casual) %>% 
  summarise(q2 = as.duration(quantile(ride_length, 0.25)),
            q3 = as.duration(quantile(ride_length, 0.5)),
            q4 = as.duration(quantile(ride_length, 0.75)),
            iqr = as.duration(IQR(ride_length)),
            lower = q2-1.5*iqr, 
            min = as.duration(round(min(ride_length))),
            upper = q4+1.5*iqr,
            max = as.duration(round(max(ride_length)))
            )

kable(df_quantile)
```


### Noting down some ideas

- Usage trends by month
- Usage trends by time of day
- Difference in ride distance

```{r}
df_filter <- df_clean %>%
  subset(select = c(ride_id, rideable_type, started_at, ended_at, ride_length, member_casual)) %>% 
  filter(as.duration(ride_length)<df_quantile$upper) %>% 
  mutate(date = as.Date(started_at)) %>% 
  mutate(year_month = ym(format(as.Date(started_at), "%Y-%m"))) %>% 
  mutate(month = month(date)) %>% 
  mutate(day = day(date)) %>% 
  mutate(year = year(date)) %>% 
  mutate(day_of_week = wday(date)) %>% 
  mutate(weekday = wday(day_of_week, label = TRUE))
```

```{r}
## Getting average amount of time different rider types ride after filtering outliers
df_stats2 <- df_filter %>% 
  group_by(member_casual) %>% 
  summarise(mean = as.duration(round(mean(ride_length))), 
            max = as.duration(round(max(ride_length))), 
            min = as.duration(round(min(ride_length))),
            sd = as.duration(round(sd(ride_length))),
            median = as.duration(round(median(ride_length)))
            )

  kable(df_stats2)
```

# Descriptive Analysis
## Ridership by type and weekday

```{r}
df_filter %>% 
  aggregate(ride_length ~ member_casual + weekday, FUN = mean) %>% 
  kable(col.names = c("Type", "Weekday", "Ride Length"))
```

## Visualizing it

```{r}
df_ride_weekday <- df_filter %>% 
  group_by(member_casual, weekday) %>% 
  summarise(rides = n(), average = mean(ride_length)) %>% 
  arrange(member_casual, weekday)

df_ride_weekday %>% 
  ggplot(aes(x = weekday, y = average/60, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Average Ride Length Between Casual Riders and Members", 
       subtitle = "Comparison by Week Day",
       y = "Average Ride Length (minutes)", 
       x = NULL, 
       fill = "Rider Type",
       caption = "Data collected between June 2021 to May 2022")
  
```
Casual riders consistently seem to rent the bikes for longer than members.

```{r}
df_ride_weekday %>% 
  ggplot(aes(x = weekday, y = rides/1000, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Comparison of Number of Rides Between Casual Riders and Members", 
       subtitle = "Comparison by Week Day",
       y = "Number of Rides (thousands)", 
       x = NULL, 
       fill = "Rider Type",
       caption = "Data collected between June 2021 to May 2022")
```

Members ride a lot more on weekdays compared to Casual riders who edge out more rides on weekends.

---

# Visualizing the number of rides per month

```{r}
df_filter %>% 
  group_by(member_casual, year_month) %>% 
  tally() %>% 
  ggplot() +
  geom_line(aes(x = year_month, y = n/1000, color = member_casual)) +
  theme_minimal() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = NULL, y = "Number of Rides (thousands)", color = "Rider Type")
```

Both types of riders seem to follow the same monthly riding trends.  Most riders are not fond of riding during winter months.

```{r}
df_filter %>%
  group_by(member_casual, year_month) %>% 
  summarise(average_ride = mean(as.numeric(as.duration(ride_length), "minutes"))) %>% 
  ggplot() +
  geom_line(aes(x = year_month, y = average_ride, color = member_casual)) +
  theme_minimal() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(limits = c(0, 30)) +
  labs(title = "Monthly Average Ride Length per Month", x = NULL, y = "Average Ride Length (minutes)", color = "Rider Type")
```

Again we can observe that casual riders tend to spend more time riding compared to members.

---

# Let's see bike types

```{r}
df_filter %>% 
  group_by(member_casual, rideable_type) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(x = rideable_type, y = count, fill = member_casual)) +
  geom_col(position = "dodge")
```

