---
title: 'Case Study: Cyclistic'
author: "Chi Kit Yeung"
date: '2022-07-22'
# output: ioslides_presentation
output: html_document
---

You will produce a report with the following deliverables:
1. A clear statement of the business task
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. A summary of your analysis
5. Supporting visualizations and key findings
6. Your top three recommendations based on your analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(janitor)
library(lubridate)
library(hms)
library(knitr)

# Loading the data
path <- "./divvy-tripdata/"
filenames <- list.files(path = path)

all_df <- lapply(filenames, function(i) {
  i <- paste(path,i,sep="")
  read.csv(i, header=FALSE)
  })
filenames <- gsub("-","_",filenames)
names(all_df) <- gsub(".csv","",filenames)

joined_df <- all_df %>% 
  reduce(full_join)

# Renaming the data frame
## reduce() function renamed the columns with some place holder values and placed the actual names into the first row of the data frame.
names(joined_df) <- joined_df[1,]
df <- joined_df %>% 
  slice(-1)

## Converting to correct data types, removing entries with NA
df_clean <- df %>% 
  mutate(started_at = ymd_hms(started_at)) %>% 
  mutate(ended_at = ymd_hms(ended_at)) %>% 
  mutate(start_lat = as.double(start_lat)) %>% 
  mutate(start_lat = as.double(start_lng)) %>%
  mutate(start_lat = as.double(end_lat)) %>%
  mutate(start_lat = as.double(end_lng)) %>%
  mutate(ride_length = as_hms(difftime(ended_at, started_at)), .after = ended_at) %>% 
  remove_empty(which = c("rows", "cols")) %>% 
  na.omit() %>% 
  filter(started_at<ended_at) %>% # remove cases where end time is somehow before or equals to the start time
  as_tibble()

df_quantile <- df_clean %>% 
  group_by(member_casual) %>% 
  summarise(q2 = as.duration(quantile(ride_length, 0.25)),
            q3 = as.duration(quantile(ride_length, 0.5)),
            q4 = as.duration(quantile(ride_length, 0.75)),
            iqr = as.duration(IQR(ride_length)),
            lower = q2-1.5*iqr, 
            min = as.duration(round(min(ride_length))),
            upper = q4+1.5*iqr,
            max = as.duration(round(max(ride_length)))
            )

df_filter <- df_clean %>%
  subset(select = c(ride_id, rideable_type, started_at, ended_at, ride_length, member_casual)) %>% 
  filter(as.duration(ride_length)<df_quantile$upper) %>% 
  mutate(date = as.Date(started_at)) %>% 
  mutate(year_month = ym(format(as.Date(started_at), "%Y-%m"))) %>% 
  mutate(month = month(date)) %>% 
  mutate(day = day(date)) %>% 
  mutate(year = year(date)) %>% 
  mutate(day_of_week = wday(date)) %>% 
  mutate(weekday = wday(day_of_week, label = TRUE))

df_ride_weekday <- df_filter %>% 
  group_by(member_casual, weekday) %>% 
  summarise(rides = n(), average = mean(ride_length)) %>% 
  arrange(member_casual, weekday)


```

## About the Company

In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that
are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and
returned to any other station in the system anytime.

## Business Task

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

## Data Source

- Documentation of steps taken to clean the data can be found in the end

```{r, echo = TRUE}
summary(cars)
```

## Executive Summary

- __Casual riders__ on average take __longer rides__ than members.
- They are also more active on __weekends__ compared to members who are active throughout the week.
- Both casual and member riders follow the same riding trends throughout the year whereriders are __least active__ during winter months.
- Although more rides used classic bikes, casual riders tend to __prefer electric bikes__ when compared to members

```{r}
df_ride_weekday %>% 
  ggplot(aes(x = weekday, y = average/60, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Average Ride Length Between Casual Riders and Members", 
       subtitle = "Comparison by Week Day",
       y = "Average Ride Length (minutes)", 
       x = NULL, 
       fill = "Rider Type",
       caption = "Data collected between June 2021 to May 2022")
```

## Visuals

```{r}
df_ride_weekday %>% 
  ggplot(aes(x = weekday, y = rides/1000, fill = member_casual)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Comparison of Number of Rides Between Casual Riders and Members", 
       subtitle = "Comparison by Week Day",
       y = "Number of Rides (thousands)", 
       x = NULL, 
       fill = "Rider Type",
       caption = "Data collected between June 2021 to May 2022")
```

## More Visuals

```{r}
df_filter %>% 
  group_by(member_casual, year_month) %>% 
  tally() %>% 
  ggplot() +
  geom_line(aes(x = year_month, y = n/1000, color = member_casual)) +
  theme_minimal() +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(x = NULL, y = "Number of Rides (thousands)", color = "Rider Type")
```

## Recommendations
- Marketing team should avoid conducting campaigns during the winter months where cycling activities is the lowest
- Marketing team should target converting casual riders during weekdays when they are most likely riding for non-recreational purposes.
- Casual riders prefer electric bikes over classic ones suggesting that they __value convenience__

